---
title: "PamDx Data Analysis Report"
params:
  author: "me"
  date: "January 1st, 1970"
  stk_qc_method: NULL
  signal_heatmap: NULL
  normalizations: NULL
  kinase_analysis: NULL
  kinase_analysis_old: NULL
  score_plot_color: NULL
  fscore_thr: NULL
  spec_thr: NULL
  psite_p_thr: NULL
  csUKA: NULL
  coral_ks_thrs: NULL
  coral_min: NULL
  coral_max: NULL
  xax_scale: NULL
  phosphosite_heatmap: NULL
  datatype: NULL
  aim: This section should be changed to reflect the aim of this experiment.
  comparisons: This section should be change to list the comparisons in this experiment.
author: "`r params$author`"
date: "`r params$date`"
always_allow_html: true
output:
  word_document:
    reference_docx: "ref/report_ref_pamdx.docx"
    toc: yes
    fig_caption: true
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(flextable)
source("R/00_GeneralFunctions.R")
source("R/01_BasicProcessing.R")
source("R/02_PhosphositeAnalysis.R")
source("R/03_KinaseAnalysis.R")
```
\pagebreak

# Introduction

The PamDx assays measure kinase activity in cell and tissue lysates, by measuring the phosphorylation of peptide representations of kinase targets/ substrates (referred hereafter as phosphosites) that are immobilized on the PamChip® microarrays. The active kinases in the sample lysates will phosphorylate their target on the array. Generic fluorescent labeled antibodies that recognize phosphorylated residues are used to visualize the phosphorylation. There are two types of PamChip® microarrays; one has 196 protein tyrosine kinase (PTK) and the other 144 serine/threonine kinase (STK) phosphosites. The supplement to this report can be found as `02_ReportSupplement_PamDx.docx`, hereafter referred to as Supplement.

## Study Information

Project aim:
`r params$aim`
Comparisons:
`r params$comparisons`

Notation used: T vs C means Test vs Control, Ratio of T/C; Values > 0 mean increased phosphorylation or kinase activity in T compared to C; Values < 0 mean lower phosphorylation or kinase activity in T compared to C. 
We recommend replicates of n=3+ per test condition for cell line experiments and n=6+ per test condition for in vivo models or primary cells. Comparisons with replicates of n<3 per group should be considered as a preliminary or trend analysis rather than biologically conclusive results. 


# Results

## Basic processing
```{r basic processing}
qc_files <- read_qc_dir()

if (nrow(qc_files) > 0) {
  qc_parsed <- parse_qc(qc_files, params$datatype)
  qc_result <- build_qc_result_text(qc_parsed$qc_table, qc_parsed$variability_table, qc_files)
}

# Auto-detected from the QC files for tercen data; falls back to the manual
# "Normalizations" UI param otherwise (bionav, or no QC files uploaded yet).
effective_normalizations <- get_effective_normalizations(qc_files, params$datatype, params$normalizations)
```

`r if(nrow(qc_files) > 0) qc_result`

```{r, echo=FALSE, results='asis'}
if (nrow(qc_files) > 0 && any(qc_files$Replicate_Type == "BR")) {
  cat("Details about QC flags and the Data Variability Indicator are provided in Supplement section 2.2.2.")
}
```

```{r, echo=FALSE, results='asis'}
norm_list <- c()
if ("vsn" %in% effective_normalizations) {
  norm_list <- c(norm_list, "VSN normalization")
} else {
  norm_list <- c(norm_list, "Log2 transformation")
}
if ("combat" %in% effective_normalizations) {
  norm_list <- c(norm_list, "ComBat correction")
}
norm_verb <- if (length(norm_list) == 1) "was" else "were"
cat(paste0("**Normalization**: ", paste(norm_list, collapse = " and "), " ", norm_verb, " applied."))

if ("vsn" %in% effective_normalizations) {
  cat("\n\nFollowing VSN normalization, observed effects are expressed relative to the overall data average.")
}
```

\newpage

## Phosphosite Analysis

```{r phosphosite analysis setup, echo=FALSE}
psite_p_thr <- if (is.null(params$psite_p_thr)) 0.05 else params$psite_p_thr
stats_files <- read_phosphosite_dir(datatype = params$datatype)

# Full names for the statistical tests, keyed by the codes in stats_files$Stats.
stats_test_labels <- c(
  Limma = "LIMMA (Linear Models for Micro Arrays)",
  MTvC  = "Multiple Test conditions versus Control (MTvC)",
  TT    = "T-test (TT)"
)
applied_tests <- unique(stats_files$Stats)
applied_tests <- applied_tests[applied_tests %in% names(stats_test_labels)]
applied_test_text <- if (length(applied_tests) == 0) {
  "statistical tests"
} else {
  paste(stats_test_labels[applied_tests], collapse = " and ")
}
```

Next, differentially phosphorylated phosphosites between conditions are identified with `r applied_test_text`.
The signal per phosphosite is the result of the phosphorylation by one or more kinases. Phosphosite-level data shows the overall trend and direction of the test effect, as well as the effect size. This is useful to place the Upstream Kinase Analysis (UKA) in context. Importantly, when the significance level in phosphosite data is low (we consider >8 significant phosphosites per assay type, per comparison as desired significance level), we consider UKA to be a preliminary or trend analysis that needs further exploration, potentially with increased number of replicates.
See also Supplement section 2.3. The table below shows the statistical analysis for each comparison.

```{r phosphosite analysis}

if (nrow(stats_files) > 0){
  stats <- parse_stats_files(stats_files, params$datatype, p_thr = psite_p_thr)
  make_report_table_stats(stats, "Phosphosite Analysis", p_thr = psite_p_thr)
}

```

Corresponding volcano plots can be found in Supplement section 2.3.

\newpage

## Upstream Kinase Analysis (UKA)

Kinases phosphorylate multiple phosphosites on the PamChip. To address which kinases are responsible for the phosphorylation differences between conditions, we perform Upstream Kinase Analysis (UKA). The UKA predicts potential kinases by examining changes in their associated phosphosite sets. The phosphosite sets are derived from publicly available databases of kinase-phosphosite relationships. For the details of the algorithm, explanation of the scores and further visualizations (kinome trees), refer to Supplement 2.4.

`r if(params$fscore_thr != 0 || params$spec_thr != 0) "The following filter(s) were used to focus on top results:"`

`r if(params$fscore_thr != 0) paste0("* ", if(params$csUKA) "Final Score" else "Median Final Score", " > ", params$fscore_thr)`
`r if(params$spec_thr != 0) paste0("* ", if(params$csUKA) "Specificity Score" else "Mean Specificity Score", " > ", params$spec_thr)`

### Top kinases table

The table below shows the top 10 kinases.

```{r kinase analysis}
kinase_files <- read_csv("temp/kinase_files.csv")

if (nrow(kinase_files) > 0) {
  kinases <- parse_kinase_files(kinase_files, fscore_thr = params$`fscore_thr`, spec_thr = params$`spec_thr`, csUKA = params$csUKA)
  make_report_table(kinases, "Kinase Analysis")
}

```

### Kinase dotplots

The Kinase dotplots below show the specificity (`r if(params$csUKA) "Specificity Score" else "Mean Specificity Score"`) and Fold Change (`r if(params$csUKA) "Kinase Statistic" else "Median Kinase Statistic"`) of the significant kinases in different comparisons. Larger dots indicate higher specificity. The kinases are clustered by families.



\newpage
# Conclusions

`r if ("vsn" %in% effective_normalizations) paste0("VSN normalization was applied therefore results reveal differences relative to the data average.")`

Explore pathways and networks with online tools. Tutorials are available [here](https://pamgene.github.io/PamGene_data_analysis_guide/PamDx%20methods/4.%20Further%20insights%20to%20PamDx%20data/). 


To validate results, refer to the guideline [here](https://pamgene.github.io/PamGene_data_analysis_guide/PamDx%20methods/4.%20Further%20insights%20to%20PamDx%20data/#suggestions-for-validating-results-with-western-blot). 

DISCLAIMER:
Support for the analysis of this project is provided for a period of 1 year. Please note that PamDx continuously improves its products and software. As a result, future projects may be conducted using updated versions, and reanalysis of existing data with newer methods may incur additional costs. In addition, the analysis of new samples originating from the same experiment at a later stage requires prior consultation and careful planning together with PamDx. Integration of such new data with the current project may be complex and could lead to additional charges.
